---
title: "Untitled"
output: html_document
date: "2024-01-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
```

Read functions for screeningmodels and screening plots
```{r}
source("code/sourcecode_screening2023-MH.R")
```

Additionally janitor is needed?

Read data and create needed variable Date from year, month and day. Keep pH, sulphate, Date and stationID (SWEREFID), rename sulphate and restructure data into long format. Run function screeningmodeling. This can take a while depending on how many series need to be processed and how complex the fitting process is. If unsure start with a subset. 
```{r}
read_tsv("data_in/Vdr 88-17 ANC300 MH.txt") %>%
  mutate(Date = paste(Yr, Mo, Day, sep="-") %>% 
           ymd())%>%
  select(`SO4* mekv/l`,pH, Date , SWEREFID)%>%
  rename(Sulfate=`SO4* mekv/l`) %>% 
  drop_na(Sulfate) %>% 
  pivot_longer(cols=c("Sulfate", "pH")) %>% 
  group_by(SWEREFID,name) %>% 
  mutate(n=length(Date)) %>% 
  screeningmodeling(values=  value,
                    datevar = Date, 
                    link = "identity", 
                    conf.type = "conf",
                    conf.level=0.95,
                    beep = TRUE, 
                    tdist = F,
                    autocor = FALSE, # FALSE ger den enklare varianten av modellen TRUE ger den mer komplexa varianten. autocor indicates whether autocorrelation should be modelled (TRUE) or not (FALSE)
                    SWEREFID, name) -> 
  screenmodel_out
```

In case the outputfile should be saved as rds file we can use the command below. This create a quite large file.
```{r}
saveRDS(screenmodel_out,"data_out/screenmodel_out_MH.rds")
#screenmodel_out <- readRDS("data_out/screenmodel_out_MH-test3.rds")
```

The created tibble screenmodel_out contains all necessary information to create the different types of plots. 

Plots for an individual series:
pH, Alsterälven (Fiugre 2, left):
```{r}
for(i in 1:length(unique(screenmodel_out$SWEREFID))) {

individual_trend <-  screenmodel_out%>%
  filter(SWEREFID==unique(screenmodel_out$SWEREFID)[i])
  
  for(j in 1:length(unique(individual_trend$name))) {

individual_trend_var <- individual_trend %>%
  filter(name==unique(individual_trend$name)[j]) %>%
    
print_individual_trend_mh(y = unique(individual_trend$name)[j])

write.table(individual_trend_var,
            paste0("data_out/individual_trend_",unique(individual_trend$name)[j],"_",unique(screenmodel_out$SWEREFID)[i],".csv"),
            sep = ",",
            fileEncoding = "utf-8",
            row.names = FALSE)

}}
  
```

Sulphate, Alsterälven (Figure 2, right)
```{r}




 plot1 <- screenmodel_out %>%
    filter(SWEREFID=="6585489-420879", name=="pH")%>%
plot_individual_trend_mh(y = "pH") %>%
ggsave("plots/plot1.png")

#write.table(plot1, 
#            sep="\t",
 #           row.names=FALSE,
   #         quote=FALSE,
     #       file ="data_out/test1.txt")
```



```{r}
screenmodel_out%>%
  filter(SWEREFID=="6585489-420879", name=="Sulfate")%>%
plot_individual_trend_mh(y = "Sulfate")
```


Trend screening plot sorted using the SWEREFID (South to north, not shown):

```{r}
screenmodel_out%>%arrange(desc(SWEREFID))%>% plot_screeningtrends(sorting = SWEREFID, y_id = SWEREFID, wrappingvar = name)
```

Trend screening plot sorted by the inverse of SWEREFID (North to South, Figure 3):
```{r}
screenmodel_out%>% mutate(order=desc(as.numeric(str_sub(SWEREFID, start=1, end=7))))%>%plot_screeningtrends(sorting = order, y_id = SWEREFID, wrappingvar = name)
```

Trend screening plot visualising p-values sorted North to South (Figure 4):
P-VÄRDEN!!!!!
```{r}
screenmodel_out%>% 
  filter(SWEREFID=="6585489-420879", name=="pH")%>%
  mutate(order=desc(as.numeric(str_sub(SWEREFID, start=1, end=7))))%>%
  print_screeningtrends_pvalues(sorting = order, y_id = SWEREFID, wrappingvar = name)
```

Proportions of increasing, decreasing and non-significant trends (Figure 5):
```{r}
screenmodel_out %>% plot_proportions(wrappingvar = name)
```



Plotting trend magnitudes compared to the mean of the first three years in each series (sorted North to South):
pH (Figure 6, upper)
```{r}
screenmodel_out %>% filter(name=="pH")%>% mutate(order=desc(as.numeric(str_sub(SWEREFID, start=1, end=7))))%>% plot_screeningtrends_reference(sorting = order, y_id = SWEREFID, wrappingvar = name)
```

Sulphate (Figure 6, lower)
```{r}
screenmodel_out %>% filter(name=="Sulfate")%>% mutate(order=desc(as.numeric(str_sub(SWEREFID, start=1, end=7))))%>% plot_screeningtrends_reference(sorting = order, y_id = SWEREFID, wrappingvar = name)
```
Adjusting the order of sites presented, e.g. by sorting accoring to reference mean (mean of the first three years, Figure 7):
```{r}
  screenmodel_out%>%
  filter(name=="Sulfate")%>%
    unnest(cols=c(predict, data))%>%
    filter(decimaldate>=decimaldate[1] & decimaldate<=decimaldate[1]+3)%>%
    summarize(reference=mean(`s(decimaldate)`+intercept, na.rm=TRUE))->refmean

screenmodel_out%>%
   filter(name=="Sulfate")%>%
  left_join(refmean)%>% 
  plot_screeningtrends_reference(sorting = reference, y_id = SWEREFID)

```

Individual series for Tore älv (Figure 10, left):
```{r}
screenmodel_out%>%
  filter(SWEREFID=="7334144-846887", name=="Sulfate")%>%
plot_individual_trend(y = "Sulfate")
```



Plotting relative trend magnitudes, i.e. change at a certain time point relative to average level at the same time point:
Sulphate (Figure 10, right):
```{r}
screenmodel_out %>% filter(name=="Sulfate")%>% mutate(order=desc(as.numeric(str_sub(SWEREFID, start=1, end=7))))%>% plot_screeningtrends_relative(sorting = order, y_id = SWEREFID, wrappingvar = name)
```




Plot data data example (not shown):
```{r}
screenmodel_out %>% plot_data %>% head
```

Exctract splines and derivatives (not shown):

```{r}
screenmodel_out$fit[[1]] %>% splines_and_derivative() %>% head()
```

```{r}
screenmodel_out$fit[[37]] %>% splines_and_derivative() %>% plot()
```

Mean values for the first three years and the last three years in each series (Supplementary, Table A1)
```{r}
screenmodel_out%>%
  filter(name=="Sulfate")%>%
    unnest(cols=c(predict, data))%>%
    filter(decimaldate<=decimaldate[n()] & decimaldate>=decimaldate[n()]-3)%>%
    summarize(averagelast=mean(`s(decimaldate)`+intercept, na.rm=TRUE))->refmeanLast

screenmodel_out %>% 
  filter(name=="Sulfate")%>%
    left_join(refmean)%>%
    left_join(refmeanLast)%>%
  arrange(reference)%>%
  select(SWEREFID, intercept, reference, averagelast)->table1
```





Referenceplot for pH (not included in paper)
```{r}
  test_out%>%
  filter(name=="pH")%>%
    unnest(cols=c(predict, data))%>%
    filter(decimaldate>=decimaldate[1] & decimaldate<=decimaldate[1]+3)%>%
    summarize(reference=mean(`s(decimaldate)`+intercept, na.rm=TRUE))->refmean

test_out%>%
   filter(name=="pH")%>%
  left_join(refmean)%>% 
  plot_screeningtrends_reference(sorting = reference, y_id = SWEREFID)
```
